### 2025-02-26 sports-cognitive-psychology.qmd
20250226 - clean up code

Adapted from Ben Baldwin: https://www.nfl4th.com/articles/4th-down-research.html

@@

nfl_4thdown %>%
  filter(vegas_wp > .2, between(go_boost, -10, 10), season %in% c(2014, 2024)) %>%
  ggplot(aes(go_boost, go, color = as.factor(season))) + 
  ggtext::geom_richtext(
    data = text_df,   
    aes(x, y, label = label, angle = angle), 
    color = "black",
    fill = NA,
    label.color = NA,
    size = 5) + 
  geom_vline(xintercept = 0) +
  stat_smooth(method = "gam", method.args = list(gamma = 1), formula = y ~ s(x, bs = "cs", k = 10), show.legend = F, se = F, linewidth = 4) +
  # this is just to get the plot to draw the full 0 to 100 range
  geom_hline(yintercept = 100, alpha = 0) +
  geom_hline(yintercept = 0, alpha = 0) +
  ggplot2::theme_classic() +
  labs(x = "Gain in win probability by going for it",
       y = "Go-for-it percentage",
       subtitle = "4th down decisions in 2024 versus 2014, win prob. > 20%",
       title = glue::glue("How <span style='color:red'>math</span> is changing football")) +
  theme(
    legend.position = "none",
    plot.title = element_markdown(size = 22, hjust = 0.5),
    plot.subtitle = element_markdown(size = 14, hjust = 0.5),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold")
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 4), limits = c(0, 100), expand = c(0,0)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10), limits = c(-10, 10), expand = c(0,0)) +
  annotate("text", x= -1.2, y= 70, label = "Should\nkick", color="black", size = 5) +
  annotate("text", x= 1.2, y= 70, label = "Should\ngo for it", color="black", size = 5) +
  geom_segment(
    aes(x = -.1, y = 80, xend = -2, yend = 80),
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black", linewidth = 2
    ) +
  geom_segment(
    aes(x = .1, y = 80, xend = 2, yend = 80),
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black", linewidth = 2
  )
```

>>>>

Adapted from Ben Baldwin: https://www.nfl4th.com/articles/4th-down-research.html (archived at <https://perma.cc/S5D8-3NCU>)

@@

nfl_4thdown %>%
  filter(vegas_wp > .2, between(go_boost, -10, 10), season %in% c(2014, 2024)) %>%
  ggplot(
    aes(
      x = go_boost,
      y = go,
      color = as.factor(season))) + 
  ggtext::geom_richtext(
    data = text_df,   
    aes(
      x,
      y,
      label = label,
      angle = angle), 
    color = "black",
    fill = NA,
    label.color = NA,
    size = 5) + 
  geom_vline(xintercept = 0) +
  stat_smooth(
    method = "gam",
    method.args = list(gamma = 1),
    formula = y ~ s(x, bs = "cs", k = 10),
    show.legend = FALSE,
    se = FALSE,
    linewidth = 4) +
  # this is just to get the plot to draw the full 0 to 100 range
  geom_hline(
    yintercept = 100,
    alpha = 0) +
  geom_hline(
    yintercept = 0,
    alpha = 0) +
  ggplot2::theme_classic() +
  labs(
    x = "Gain in win probability by going for it",
    y = "Go-for-it percentage",
    subtitle = "4th down decisions in 2024 versus 2014, win prob. > 20%",
    title = glue::glue("How <span style='color:red'>math</span> is changing football")) +
  theme(
    legend.position = "none",
    plot.title = element_markdown(size = 22, hjust = 0.5),
    plot.subtitle = element_markdown(size = 14, hjust = 0.5),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold")
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 4),
    limits = c(0, 100),
    expand = c(0,0)) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 10),
    limits = c(-10, 10),
    expand = c(0,0)) +
  annotate(
    "text",
    x = -1.2,
    y = 70,
    label = "Should\nkick",
    color = "black",
    size = 5) +
  annotate(
    "text",
    x = 1.2,
    y = 70,
    label = "Should\ngo for it",
    color = "black",
    size = 5) +
  geom_segment(
    aes(
      x = -.1,
      y = 80,
      xend = -2,
      yend = 80),
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black",
    linewidth = 2) +
  geom_segment(
    aes(
      x = .1,
      y = 80,
      xend = 2,
      yend = 80),
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black",
    linewidth = 2)
```
